Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-29 paper.tex
Text tweaks

However, what about a simulation with changing population size?
To understand how runtime depends on demography
it helps to consider why runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

Putting this together:
suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$ -- in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
In practice, we recommend that users estimate runtime by
measuring runtime in a series of simulations with short genome lengths,
possibly varying other parameters to check for interactions,
and then predict runtime by fitting a quadratic curve to genome length.
There is only a weak dependency on sample size
-- notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
only increases runtime by a factor of 2 or so.

an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}, to facilitate
simulation of organisms such as \textit{Drosophila melanogaster}.

>>>>

However, what about a simulation with changing population size?
To understand how runtime depends on demography
it helps to consider why runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that
lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal.
Threfore, many (roughly half) of these lineages have at
least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
The number of linages decreases gradually,
on the coalescent time scale---so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens is roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$, as desired.

What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively
evenly on the coalescent time scale: in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.
Suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
In practice, we recommend that users estimate runtime by
measuring runtime in a series of simulations with short genome lengths,
possibly varying other parameters to check for interactions,
and then predict runtime by fitting a quadratic curve to genome length.
There is only a weak dependency on sample size
---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
only increases runtime by a factor of 2 or so.
\jkcomment{as expected, since it's $O(\log^2{n})$?}
